\documentclass[11pt, letterpaper]{article}

% Essential Packages
\usepackage{amsmath, amssymb, amsthm}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{bm}
\usepackage{microtype}

% Page Margins
\geometry{left=1in, right=1in, top=1in, bottom=1in}

% Theorem Styles
\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

% Custom Math Macros (Example)
\newcommand{\R}{\mathbb{R}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\Q}{\mathbb{Q}}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\diag}{diag}

\begin{document}

\section*{Phase 6.2 — Heteroscedastic GP Surrogate}

\subsection*{Feature map}

Let $\mathcal{P}$ denote the space of \textsc{Package} configurations.
A package $p \in \mathcal{P}$ is encoded as a fixed-length feature vector

\[
  \bm{x} = \varphi(p) \in \mathbb{R}^{d}
\]

where $\varphi$ concatenates one-hot encodings of the model slot, a binary
skill-presence vector, and one-hot encodings of the named system-prompt and
template-value variants (the Phase~3.1 slot-space schema).

\subsection*{Observations}

After $N$ completed trials we have a dataset
$\mathcal{D}_N = \{(\bm{x}_i, \bm{y}_i)\}_{i=1}^{N}$ where each
$\bm{y}_i \in \mathbb{R}^{5}$ records the five Pareto-axis values

\[
  \bm{y}_i =
  \bigl(
    \bar{\tau}_i,\;
    \bar{c}_i,\;
    s_i,\;
    \bar{q}_i,\;
    r_i
  \bigr)
  \qquad
  \text{(tokens, dollars, scaling slope, quality, subjective)}
\]

Trials without a subjective score contribute only the first four components;
the subjective head is trained on the subset of fully-scored trials.

\subsection*{Heteroscedastic single-task GP (per output head)}

For each output $j \in \{1,\ldots,5\}$ we place an independent GP prior:

\[
  f^{(j)}(\bm{x}) \sim \mathcal{GP}\!\left(0,\; k^{(j)}(\bm{x}, \bm{x}')\right)
\]

with a squared-exponential (RBF-ARD) kernel

\[
  k^{(j)}(\bm{x}, \bm{x}') =
  \sigma_{f,j}^{2} \exp\!\left(
    -\tfrac{1}{2}(\bm{x} - \bm{x}')^{\top}
    \bm{\Lambda}_j^{-1}
    (\bm{x} - \bm{x}')
  \right)
\]

where $\bm{\Lambda}_j = \operatorname{diag}(\ell_{j,1}^{2}, \ldots, \ell_{j,d}^{2})$
are per-dimension length-scales (automatic relevance determination).

\paragraph{Input-dependent noise.}
Rather than assuming a fixed noise level, we model the log noise variance
with a second GP:

\[
  g^{(j)}(\bm{x}) \sim \mathcal{GP}\!\left(0,\; k_{\varepsilon}^{(j)}(\bm{x},\bm{x}')\right)
  \qquad
  \sigma^{2,(j)}(\bm{x}) = \exp\!\left(g^{(j)}(\bm{x})\right)
\]

The observation model is then

\[
  y_i^{(j)} = f^{(j)}(\bm{x}_i) + \varepsilon_i^{(j)},
  \qquad
  \varepsilon_i^{(j)} \sim \mathcal{N}\!\left(0,\; \sigma^{2,(j)}(\bm{x}_i)\right)
\]

\subsection*{Posterior}

Let $\bm{X} = [\bm{x}_1,\ldots,\bm{x}_N]^{\top}$,
$\bm{y}^{(j)} = [y_1^{(j)},\ldots,y_N^{(j)}]^{\top}$, and
$\bm{\Sigma}^{(j)} = \operatorname{diag}\!\left(\sigma^{2,(j)}(\bm{x}_1),\ldots,\sigma^{2,(j)}(\bm{x}_N)\right)$.

The posterior over $f^{(j)}$ at a new point $\bm{x}_*$ is Gaussian:

\begin{align}
  \mu_N^{(j)}(\bm{x}_*)
    &= \bm{k}_*^{(j)\top}
       \left[\bm{K}^{(j)} + \bm{\Sigma}^{(j)}\right]^{-1}
       \bm{y}^{(j)} \\[4pt]
  \sigma_N^{2,(j)}(\bm{x}_*)
    &= k^{(j)}(\bm{x}_*,\bm{x}_*)
      - \bm{k}_*^{(j)\top}
        \left[\bm{K}^{(j)} + \bm{\Sigma}^{(j)}\right]^{-1}
        \bm{k}_*^{(j)}
\end{align}

where $\bm{K}^{(j)}_{ij} = k^{(j)}(\bm{x}_i, \bm{x}_j)$ and
$\bm{k}_*^{(j)} = [k^{(j)}(\bm{x}_*, \bm{x}_i)]_{i=1}^N$.

Hyperparameters
$\theta_j = \{\sigma_{f,j}^2,\, \bm{\Lambda}_j,\, \theta_{\varepsilon,j}\}$
are optimised by maximising the log marginal likelihood.

\subsection*{Bootstrap discipline}

Define the trial count threshold $N_0 \approx 10$.  The proposer operates
in two regimes:

\[
  \text{next proposal} =
  \begin{cases}
    \texttt{RandomFromSlotSpace} & N < N_0 \\[4pt]
    \argmax_{\bm{x} \in \mathcal{X}}
      \;\alpha_{\mathrm{EHVI}}\!\left(\bm{x};\,\mathcal{D}_N\right)
    & N \geq N_0
  \end{cases}
\]

where $\mathcal{X}$ is the discrete set of un-evaluated packages and
$\alpha_{\mathrm{EHVI}}$ is the expected hypervolume improvement over the
current 5D Pareto frontier (Phase~6.3).

\subsection*{Note}

A few design notes embedded in the math:

\begin{itemize}
  \item \textbf{ARD kernel:} length-scales $\ell_{j,k}$ let the model learn which slots
  actually matter per axis — useful since model is likely far more
  predictive of dollars than \texttt{system\_prompt}.
  \item \textbf{Independent heads:} simpler than a multi-task GP and sufficient
  here; cross-axis correlations (e.g. tokens $\leftrightarrow$ dollars) can be
  captured later if needed.
  \item \textbf{Heteroscedastic noise:} the subjective head in particular will have
  high variance at low trial counts and should narrow as more scores
  arrive — homoscedastic noise would mask that. 
  \item \textbf{Bootstrap at $N_0 \approx 10$:} the GP is unreliable below this; pure random
  exploration avoids over-exploiting a badly-conditioned posterior.
\end{itemize}
\end{document}
